Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-06 talk/dls2012/paper.tex
replace some instances of "PyPy" with "RPython"

\section{Background: PyPy}
\label{sec:PyPy}

@@

PyPy's tracing JIT compiler traces on the level of RPython programs. Thus it
actually traces the execution of an interpreter written in RPython, not of the
program itself. This makes the details of the object model of the implemented
language transparent and optimizable by the tracing JIT. In the context of this
paper, this aspect of PyPy's tracing JIT can be ignored. Instead, it is
sufficient to view PyPy's tracing JIT as a JIT for RPython.

@@

The first line is a label $L_0$ with argument $i_0$. Every label has a list of
arguments. The \lstinline{print} operation just prints its argument (it is not
an operation that PyPy's tracing JIT really supports, we just use it for this
example). The \lstinline{jump} operation jumps back to the beginning of the
trace, listing the new values of the arguments of the trace. In this case, the
new value of $i_0$ is $i_0$, making it a loop-invariant.

@@

If a pure operation appears more than once in the trace with the same input
arguments, it only needs be executed the first time and then the result
can be reused for all other appearances. PyPy's optimizers can also remove
repeated heap reads if the intermediate operations cannot have changed their
value.\footnote{We perform a type-based alias analysis to know which
writes can affect which reads~\cite{XXX}. In addition writes on newly allocated objects
can never change the value of old existing ones.}

@@

PyPy's allocation removal optimization~\cite{bolz_allocation_2011} makes it
possible to identify objects that are allocated within the loop but never
escape it. That is, no outside
object ever gets a reference to them. This
is performed by processing the operations in order and
optimistically removing every \lstinline{new} operation. Later on if
it is discovered that a reference to the object escapes the loop, the
\lstinline{new} operation is inserted at this point. All operations
(\lstinline{get}, \lstinline{set} and \lstinline{guard}) on the removed objects
are also removed and the optimizer needs to keep track of the value of all used
attributes of the object.

@@

The loop peeling optimization was implemented in the PyPy
framework in about 450 lines of RPython code. That means that the JIT-compilers generated for all
interpreters implemented within PyPy now can take advantage of
it. Benchmarks have been executed for a few different interpreters and
we see improvements in several cases. The ideal loop for this optimization
is short and contains numerical calculations with no failing guards and no
external calls. Larger loops involving many operations on complex objects
typically benefit less from it. Loop peeling never makes runtime performance worse, in
the worst case the peeled loop is exactly the same as the preamble. Therefore we
chose to present benchmarks of small numeric kernels where loop peeling can show
its use.

@@

\subsection{Python}
The Python interpreter of the PyPy framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python and in C and their
runtimes are compared in Figure~\ref{fig:benchmarks}. The benchmarks are
\begin{itemize}
\item {\bf sqrt}: approximates the square root of $y$ as $x_\infty$
  with $x_0=y/2$ and $x_k = \left( x_{k-1} + y/x_{k-1} \right) /
  2$. There are three different versions of this benchmark where $x_k$
  is represented with different type of objects: int's, float's and
  Fix16's. The latter, Fix16, is a custom class that implements
  fixpoint arithmetic with 16 bits precision. In Python there is only
  a single implementation of the benchmark that gets specialized
  depending on the class of it's input argument, $y$, while in C,
  there are three different implementations.
\item {\bf conv3}: one-dimensional convolution with fixed kernel-size $3$.
\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$.
\item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two-dimensional
  arrays.
\item {\bf dilate3x3}: two-dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the elements, the maximum is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}: a low-level video processing algorithm used to
  locate edges in an image. It calculates the gradient magnitude
  using sobel derivatives. 
\end{itemize}

@@

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in previous
work~\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
speedup of loop peeling is 70\%, which makes benchmark times
comparable with native-compiled C code. We attribute the performance gap to C code to
the relative immaturity of PyPy's JIT assembler backend as well as missing
optimizations, like instruction scheduling.

>>>>

\section{Background: RPython and PyPy}
\label{sec:PyPy}

@@

RPython's tracing JIT compiler traces on the level of RPython programs. Thus it
actually traces the execution of an interpreter written in RPython, not of the
program itself. This makes the details of the object model of the implemented
language transparent and optimizable by the tracing JIT. In the context of this
paper, this aspect of RPython's tracing JIT can be ignored. Instead, it is
sufficient to view RPython's tracing JIT as a JIT for RPython.

@@

The first line is a label $L_0$ with argument $i_0$. Every label has a list of
arguments. The \lstinline{print} operation just prints its argument (it is not
an operation that RPython's tracing JIT really supports, we just use it for this
example). The \lstinline{jump} operation jumps back to the beginning of the
trace, listing the new values of the arguments of the trace. In this case, the
new value of $i_0$ is $i_0$, making it a loop-invariant.

@@

If a pure operation appears more than once in the trace with the same input
arguments, it only needs be executed the first time and then the result
can be reused for all other appearances. RPython's optimizers can also remove
repeated heap reads if the intermediate operations cannot have changed their
value.\footnote{We perform a type-based alias analysis to know which
writes can affect which reads~\cite{XXX}. In addition writes on newly allocated objects
can never change the value of old existing ones.}

@@

RPython's allocation removal optimization~\cite{bolz_allocation_2011} makes it
possible to identify objects that are allocated within the loop but never
escape it. That is, no outside
object ever gets a reference to them. This
is performed by processing the operations in order and
optimistically removing every \lstinline{new} operation. Later on if
it is discovered that a reference to the object escapes the loop, the
\lstinline{new} operation is inserted at this point. All operations
(\lstinline{get}, \lstinline{set} and \lstinline{guard}) on the removed objects
are also removed and the optimizer needs to keep track of the value of all used
attributes of the object.

@@

The loop peeling optimization was implemented in the PyPy
framework in about 450 lines of RPython code. That means that the JIT-compilers generated for all
interpreters implemented with RPython now can take advantage of
it. Benchmarks have been executed for a few different interpreters and
we see improvements in several cases. The ideal loop for this optimization
is short and contains numerical calculations with no failing guards and no
external calls. Larger loops involving many operations on complex objects
typically benefit less from it. Loop peeling never makes runtime performance worse, in
the worst case the peeled loop is exactly the same as the preamble. Therefore we
chose to present benchmarks of small numeric kernels where loop peeling can show
its use.

@@

\subsection{Python}
The Python interpreter of the RPython framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python and in C and their
runtimes are compared in Figure~\ref{fig:benchmarks}. The benchmarks are
\begin{itemize}
\item {\bf sqrt}: approximates the square root of $y$ as $x_\infty$
  with $x_0=y/2$ and $x_k = \left( x_{k-1} + y/x_{k-1} \right) /
  2$. There are three different versions of this benchmark where $x_k$
  is represented with different type of objects: int's, float's and
  Fix16's. The latter, Fix16, is a custom class that implements
  fixpoint arithmetic with 16 bits precision. In Python there is only
  a single implementation of the benchmark that gets specialized
  depending on the class of it's input argument, $y$, while in C,
  there are three different implementations.
\item {\bf conv3}: one-dimensional convolution with fixed kernel-size $3$.
\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$.
\item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two-dimensional
  arrays.
\item {\bf dilate3x3}: two-dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the elements, the maximum is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}: a low-level video processing algorithm used to
  locate edges in an image. It calculates the gradient magnitude
  using sobel derivatives. 
\end{itemize}

@@

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in previous
work~\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
speedup of loop peeling is 70\%, which makes benchmark times
comparable with native-compiled C code. We attribute the performance gap to C code to
the relative immaturity of RPython's JIT assembler backend as well as missing
optimizations, like instruction scheduling.
